---
author: "Prashant Kumar"
format: 
  html:
    code-fold: true
    toc: true
    toc-location: right
    embed-resources: true
editor: source
---
# <span style="color:blue;">DNA sequence Generation Prediction</span>

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(readxl)
library(dplyr)
library(stringr)
library(gt)
library(ggplot2)
library(plotly)
library(openxlsx)
library(tidyr)
library(scales)
library(pROC)
```
## Data Cleaning
```{r message=FALSE, warning=FALSE, echo=FALSE}
data <- read.csv("E:/micobiome core center uiowa analysis/project 1/ECMP NextSeq Run Metrics_2.csv", check.names = FALSE) %>% 
  .[, -11] %>%  
  rename(sample_number = `Sample Number`,
         next_seq_run = `NextSeq Run #/ Date`,
         library_number = `Library No. / Date`,
         panel_name = `Panel name`,
         `loding_conc(pM)` = `Loading Conc. (pM)`,
         cluster_density = `Cluster Density (K/mm2)`,
         `G%` = `(G) % > = Q30`,
         library_conc_nM = `Lib. Conc (nM)`,
         library_dilution_2nM = `Library dilution to 2nM`,
         AmpliSeq_Index = `AmpliSeq Index`,
         Patient_Gender = `Patient Gender`,
         Sample_ID = `Sample ID`,
         Total_PF_Reads = `Total PF Reads`,
         Amplicon_Mean_Coverage = `Amplicon Mean Coverage`,
         Uniformity_coverage = `Uniformity of coverage (%)`,
         Percent_Ontarget_Aligned_Reads = `Percent On-target Aligned Reads`,
         `Total_Aligned_reads_persample(R1/R2)` =`Total Aligned reads per sample R1/R2`,
         `Percent_Aligned_Bases(R1/R2)` = `Percent Aligned Bases R1/R2`) %>% 
  rename_all(tolower) %>% 
  select(sample_number, library_conc_nm, total_pf_reads,
         amplicon_mean_coverage, cluster_density, 
         uniformity_coverage) %>%
    mutate(
    total_pf_reads = gsub(",", "", total_pf_reads),
    total_pf_reads = ifelse(
      total_pf_reads %in% c(" ","","Not Sequenced", "NA", 
                            "not Sequenced", 
                            "CTS Decalcified", 
                            "not sequenced", 
                            "***failed library, repeat next run***",
                            "***failed library repeat next run***"),
      NA, total_pf_reads),
    amplicon_mean_coverage = ifelse(
      amplicon_mean_coverage %in% c(" ","","Not Sequenced", "NA", 
                                    "not Sequenced",
                                    "CTS Decalcified", "not sequenced", 
                                    "**Failed library prep, column clean and repeat next run**"), 
                                    NA, amplicon_mean_coverage), 
    amplicon_mean_coverage = ifelse(
      amplicon_mean_coverage == "87.43.5", "8743.5", 
      amplicon_mean_coverage),
    uniformity_coverage = as.numeric(gsub("%", "", uniformity_coverage)),
    cluster_density = gsub("\\s*\\+/-.*", "", cluster_density),
    cluster_density = as.numeric(cluster_density)
  ) %>% 
    filter(
    !library_conc_nm %in% c("", " ",
                            "Removed",
                            "not sequenced",
                            "no dilution",
                            "run straight",
                            "cc and repeat",
                            "FAIL", "FAIL ",
                            "Failed library prep, removed from run",
                            "NA")) %>% 
  mutate(across(where(is.character), ~ na_if(., ""))) %>% 
  mutate(across(everything(), as.numeric),
         sequence = as.factor(case_when(
           is.na(amplicon_mean_coverage) ~ "No",
           amplicon_mean_coverage > 1000 & uniformity_coverage <= 75 ~ "QNS",
           TRUE ~ "Yes")
           )) %>% 
  filter(!is.na(library_conc_nm),library_conc_nm <= 100,
         (is.na(amplicon_mean_coverage) | amplicon_mean_coverage >= 1000)) %>% 
  mutate(generated = factor(ifelse(sequence %in% c("No", "QNS"), 0, 1)),
         uniformity_coverage = ifelse(is.na(uniformity_coverage), 0, uniformity_coverage),
         generated = factor(generated, levels = c(0, 1), labels = c("No", "Yes"))
  )

nextseq <- data %>% 
  mutate(group = cumsum(!is.na(cluster_density))) %>% 
  group_by(group) %>% 
  mutate(sample_number = row_number()) %>% 
  fill(cluster_density, .direction = "downup") %>% 
  ungroup()  %>% 
  select(-group) 


        
```

## NA Distribution Or not Sequenced data.
```{r message=FALSE, warning=FALSE, echo=FALSE}
na_distribution <- nextseq %>%
  group_by(library_conc_nm) %>%
  summarise(across(c(-sequence, -generated), ~ sum(is.na(.))))
# head(na_distribution)

counts <- nextseq %>%
  group_by(sequence) %>%
  mutate(
    sequence = factor(sequence, 
                           levels = c("Yes", "No", "QNS"))
    ) %>% 
  summarise(count = n()) 
```


## Prediction of Whether sequence is generated or not based on Library concentration.
```{r message=FALSE, warning=FALSE, echo=FALSE}
model <- glm(generated ~ library_conc_nm, data = nextseq, family = binomial)
summary(model)
predicted_probs <- predict(model, type = "response")
roc_curve <- roc(nextseq$generated, predicted_probs)
exp(coef(model))
plot(roc_curve)
auc(roc_curve)
```

1) The log-odds of the sequence is generated when library concentration is 0 is 0.96779, that means model predicts a 72.46% probability of sequence generation at baseline. \n
2) Log-Odds at library Concentration = 1 is (0.96779 + 0.28903). \n
That means odds = 4.88, and model predict a 77.84% probability of sequence being generated at library Concentration = 1. \n
3) Every one unit increase in the library concentration lead the probability of increase in the sequence generation by 5.39%. \n
4) P value = 2e-16, suggesting library concentration is a significant predictor of whether sequence is generated or not. \n


## Model to check, Is total_pf_reads Missing Values Based on library Concentration.
```{r message=FALSE, warning=FALSE, echo=FALSE}
model1 <- glm(generated ~ library_conc_nm + cluster_density, data = nextseq, family = binomial)
summary(model1)
predicted_probs <- predict(model1, type = "response")
roc_curve <- roc(nextseq$generated, predicted_probs)
exp(coef(model1))
plot(roc_curve)
auc(roc_curve)
```


## Count of Sequence Generation by Library concentration and Cluster Density.
```{r message=FALSE, warning=FALSE, echo=FALSE}
nextseq %>% 
  filter(library_conc_nm <= 10) %>%
  mutate(
    sequence = factor(sequence, 
                           levels = c("Yes", "No", "QNS"))
         ) %>% 
  ggplot(aes(x = library_conc_nm, fill = sequence)) +
  geom_bar(position = "stack") +
  scale_x_continuous(minor_breaks = seq(0, 10, by = 0.5),
                     breaks = seq(0, 10, by = 1), 
                     labels = label_number()) + 
  scale_fill_manual(values = c("Yes" = "#fa1e75", 
                               "No" = "#1f77b4", 
                               "QNS" = "#2ca02c"))  +
  labs(title = "Sequence Generation by Library Concentration",
       x = "Library Concentration (nM)",
       y = "Count of Sequence Generation",
       fill = "Sequence Generated") +
  theme_classic() +
  theme(axis.text.x = element_text(hjust = 1),
        legend.position = "none")


nextseq %>% 
  filter(library_conc_nm <= 10 & library_conc_nm > 0) %>% 
  mutate(
    sequence = factor(sequence, levels = c("Yes", "No", "QNS"))
  ) %>% 
  group_by(library_conc_nm, sequence) %>% 
  summarise(count = n(), .groups = 'drop') %>%
  ggplot(aes(x = library_conc_nm, y = count, color = sequence)) +
  geom_line(aes(group = sequence), size = 0.6, linetype = "solid") +
  scale_color_manual(values = c("Yes" = "#fa1e75", 
                               "No" = "#1f77b4", 
                               "QNS" = "#2ca02c")) +
  scale_x_continuous(breaks = seq(0, 10, by = 1)) +
  labs(
    title = "Sequence Generation by Library Concentration",
    x = "Library Concentration (nM)",
    y = "Count of Sequence Generation", 
    fill = "Sequence Generated"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(hjust = 0.5),
        legend.position = "none")

nextseq %>%
  filter(cluster_density < 400) %>%
  mutate(
    sequence = factor(sequence, levels = c("Yes", "No", "QNS"))
  ) %>%
  group_by(cluster_density, sequence) %>%
  summarise(count = n(), .groups = 'drop') %>%
  ggplot(aes(x = cluster_density, y = count, color = sequence)) +
  geom_line(aes(group = sequence), size = 0.6, linetype = "solid") + 
  scale_x_continuous(limits = c(50, 400),
                     minor_breaks = seq(50, 400, by = 10),
                     breaks = seq(50, 400, by = 50)) +
   scale_y_continuous(limits = c(0, 100),
                      minor_breaks = seq(0, 100, by = 2),
                      breaks = seq(0, 100, by = 10)) + 
  scale_color_manual(values = c("Yes" = "#fa1e75", 
                               "No" = "#1f77b4", 
                               "QNS" = "#2ca02c")) +
  labs(title = "Sequence Generation by Cluster Density",
       x = "Cluster Density (K/mm²)",
       y = "Number of Samples",
       color = "Sequence Generated") +
  theme_classic() +
  theme(
    axis.ticks.y = element_line(size = 0.1, color = "gray"),
    axis.ticks.length = unit(0.1, "cm"), 
    axis.ticks.major = element_line(size = 0.1, color = "gray"),
    axis.ticks.minor = element_line(size = 0.1, color = "gray"),
    panel.grid.minor = element_line(color = "gray", size = 0.2,),
    panel.grid.major = element_line(color = "gray", size = 0.2), 
    panel.background = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 0, vjust = 0.5)
  ) 

# ggsave("E:/micobione core center uiowa analysis/project 1/fig10.tiff", width = 30, height = 10)

```


## Boxplot of Library Concentration and Cluster Dendity by Sequence
```{r message=FALSE, warning=FALSE, echo=FALSE}
nextseq %>% 
  filter(library_conc_nm <= 30) %>%
  mutate(
    sequence = factor(sequence, levels = c("Yes", "No", "QNS"))
  ) %>% 
  ggplot(aes(x = sequence, 
           y = library_conc_nm, 
           fill = sequence)) +
  geom_boxplot() +
  labs(title = "Library Concentration Distribution by Sequence Generation",
       x = "Sequence Generation",
       y = "Library Concentration (nM)") +
  scale_y_continuous(minor_breaks = seq(0, 30, by = 1),
                     breaks = seq(0, 30, by = 5), 
                     labels = label_number()) +
  scale_fill_manual(values = c("Yes" = "#fa1e75", 
                               "No" = "#1f77b4", 
                               "QNS" = "#2ca02c")) +
  theme(
    axis.ticks.y = element_line(size = 0.25, color = "gray"),
    axis.ticks.length = unit(0.1, "cm"), 
    axis.ticks.major = element_line(size = 0.25, color = "gray"),
    axis.ticks.minor = element_line(size = 0.25, color = "gray"),
    panel.grid.minor = element_line(color = "gray", size = 0.2),
    panel.grid.major = element_line(color = "gray", size = 0.5),
    legend.position = "none"
  ) +
theme(panel.background = element_rect(fill = "white")) +
  stat_summary(fun = median, 
               geom = "text", 
               aes(label = round(..y.., 2)), 
               position = position_nudge(y = 1),
               color = "black", size = 3) +
geom_text(data = counts, 
            aes(x = sequence,y = c(15, 5, 11), 
                label = paste("n =", count)), 
          color = "black", size = 4, fontface = "bold", vjust = -0.5)


nextseq %>% 
  filter(sequence %in% c("Yes", "QNS")) %>% 
  ggplot(aes(x = amplicon_mean_coverage, 
           y = sequence, color = sequence)) +
  geom_point() +  
  geom_jitter(width = 0.5, alpha = 0.2)+
  scale_fill_manual(values = c("Yes" = "#fa1e75", "QNS" = "#2ca02c")) +
  labs(
    title = "Amplicon Mean Coverage by Sequence Outcome",
    x = "Amplicon Mean Coverage",
    y = "Sequence Outcome"
  ) + 
    scale_x_continuous(limits = c(0, 40000),
      minor_breaks = seq(0, 40000, by = 2500),
                     breaks = seq(0, 40000, by = 5000)) +
  theme_classic()+
  theme(
    panel.grid.minor = element_line(color = "gray", size = 0.25),
    axis.ticks.x = element_line(size = 0.25, color = "gray"),
    axis.ticks.length = unit(0.2, "cm"),  # Length of the ticks
    axis.ticks.major = element_line(size = 0.25, color = "gray"),
    axis.ticks.minor = element_line(size = 0.25, color = "gray")
  )

nextseq %>% filter(cluster_density < 400) %>% 
  mutate(
    sequence = factor(sequence, 
                           levels = c("Yes", "No", "QNS"))
         ) %>%  
  ggplot(aes(x = sequence, 
           y = cluster_density, 
           fill = sequence)) +
  geom_boxplot() +
  labs(title = "Cluster Density Distribution by Sequence Generation",
       x = "Sequence Generation",
       y = "Cluster Density (K/mm2)") +
  scale_y_continuous(minor_breaks = seq(50, 400, by = 10),
                     breaks = seq(50, 400, by = 50), 
                     labels = label_number()) +
  scale_fill_manual(values = c("Yes" = "#fa1e75", 
                               "No" = "#1f77b4", 
                               "QNS" = "#2ca02c")) +
  theme(
    axis.ticks.y = element_line(size = 0.25, color = "gray"),
    axis.ticks.length = unit(0.1, "cm"), 
    axis.ticks.major = element_line(c),
    axis.ticks.minor = element_line(size = 0.25, color = "gray"),
    panel.grid.minor = element_line(color = "gray", size = 0.2),
    panel.grid.major = element_line(color = "gray", size = 0.5),
    legend.position = "none"
  ) +
theme(panel.background = element_rect(fill = "white")) +
  stat_summary(fun = median, 
               geom = "text", 
               aes(label = round(..y.., 2)), 
               position = position_nudge(y = 1),
               color = "black", size = 3) +
geom_text(data = counts, 
            aes(x = sequence,y = c(240, 240, 235), 
                label = paste("n =", count)), 
          color = "black", size = 4, fontface = "bold", vjust = -0.5)

```

## Density of Sequence generation by Library concentration.
```{r message=FALSE, warning=FALSE, echo=FALSE}
nextseq %>% 
  filter(library_conc_nm <= 20) %>% 
    mutate(sequence = factor(sequence, 
                           levels = c("Yes", "No", "QNS"))
         ) %>%
  ggplot(aes(x = library_conc_nm, fill = sequence)) +
  geom_density(alpha = 0.6) +
  labs(title = "Density of Sequence Generation by Library Concentration",
       x = "Library Concentration (nM)",
       y = "Density",
       fill = "Sequence Generated") + 
  geom_text(data = counts, 
            aes(x = c(8, 0.5, 3.5), y = c(0.08, 0.24, 0.09), 
                label = paste("n =", count)), 
            color = "black", size = 4, fontface = "bold", vjust = -0.5) +
  scale_x_continuous(breaks = seq(0, 20, by = 2))+
  theme_classic() +
  scale_fill_manual(values = c("Yes" = "#fa1e75", "No" = "#1f77b4", "QNS" = "#2ca02c"))

nextseq %>% 
  filter(cluster_density < 400) %>% 
    mutate(sequence = factor(sequence, 
                           levels = c("Yes", "No", "QNS"))
         ) %>%
  ggplot(aes(x = cluster_density, fill = sequence)) +
  geom_density(alpha = 0.6) +
  labs(title = "Density of Sequence Generation by Cluster Density",
       x = "Cluster Density",
       y = "Density",
       fill = "Sequence Generated") + 
  geom_text(data = counts,
            aes(x = c(235, 170, 140), y = c(0.0085, 0.0095, 0.0065),
                label = paste("n =", count)),
            color = "black", size = 4, fontface = "bold", vjust = -0.5) +
  scale_x_continuous(breaks = seq(50, 400, by = 50))+
  theme_classic() +
  scale_fill_manual(values = c("Yes" = "#fa1e75", "No" = "#1f77b4", "QNS" = "#2ca02c"))
```

# count of sequence between cluster density between 150 to 230 (K/mm2).
```{r}
nextseq %>% 
  filter(cluster_density >= 150 & cluster_density<= 230) %>%
  group_by(sequence) %>%
  mutate(
    sequence = factor(sequence, 
                           levels = c("Yes", "No", "QNS"))
    ) %>% 
  summarise(count = n()) %>% 
  gt() %>% 
    tab_header(
    title = "count of sequence between cluster density between 150 to 230 (K/mm2)"
  ) 
```

# count of sequence between Library concentration between 2 to 5 (nm).
```{r, echo=FALSE}
nextseq %>% 
  filter(library_conc_nm >= 2 & library_conc_nm <= 5) %>%
  group_by(sequence) %>%
  mutate(
    sequence = factor(sequence, 
                           levels = c("Yes", "No", "QNS"))
    ) %>% 
  summarise(count = n()) %>% 
  gt() %>% 
    tab_header(
    title = "Count of Sequence between Library Concentration 2 to 5 (nm)"
  ) 
```


```{r message=FALSE, warning=FALSE, include=FALSE}
nextseq %>%
  select(library_conc_nm, total_pf_reads, amplicon_mean_coverage, sequence) %>% 
  head() %>%
  gt() %>%
  cols_label(
    library_conc_nm = "Library Concentration",
    total_pf_reads = "Total PF Reads",
    amplicon_mean_coverage = "Amplicon Mean Coverage",
    sequence = "Sequence Generated"
  ) %>%
  tab_header(
    title = md("**NextSeq Data Summary**"),
    subtitle = md("*Showing the first few rows*")
  ) %>% 
  tab_spanner(
    label = "Library Information",
    columns = c(library_conc_nm)
  ) %>%
  tab_spanner(
    label = "Reads and Coverage",
    columns = c(total_pf_reads, amplicon_mean_coverage)
  ) %>%
  tab_spanner(
    label = "Sequence Information",
    columns = c(sequence)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold", color = "darkblue"),
    locations = cells_title()
  ) %>%
 tab_style(
    style = cell_text(color = "#25dc00"),
    locations = cells_body(rows = sequence == "Yes", columns = sequence)
  ) %>%
  tab_style(
    style = cell_text(color = "#ff5d12"),
    locations = cells_body(rows = sequence == "No", columns = sequence)
  )
```

## Prediction Model
Prediction Model for the probability of sequence generation based on the library concentration of the previous model.
$\text{log-odds} = \beta_0 + \beta_1 \times \text{library concentration} + \beta_2 \times \text{cluster density}$ <br>
$\text{Probability} = 1/(1 + e^{-\text{log-odds}})$ <br>
$\text{Probability} = 1/(1 + e^{-(\beta_0 + \beta_1 \times \text{library concentration} + \beta_2 \times \text{cluster density})})$ <br>

```{r message=FALSE, warning=FALSE, include=FALSE}
set.seed(9999)
model <- glm(generated ~ library_conc_nm + cluster_density, data = nextseq, family = binomial)

prediction_sequence_generation <- function(library_conc_nm, cluster_density, model) {
  intercept <- model$coefficients[1]
  coef_library_conc <- model$coefficients[2]
  coef_cluster_density <- model$coefficients[3]
  
  log_odds <- intercept + coef_library_conc * library_conc_nm + coef_cluster_density * cluster_density
  
  probability <- 1 / (1 + exp(-log_odds))
  
  return(probability)
}

library_conc_values <- runif(1000, min = 0, max = 30) 
uniformity_coverage_values <- runif(100, min = 0, max = 100) 

predicted_probabilities <- mapply(prediction_sequence_generation, 
                                  library_conc_values, 
                                  uniformity_coverage_values, 
                                  MoreArgs = list(model = model))

predicted_data <- data.frame(library_concentration = library_conc_values,
                             uniformity_coverage = uniformity_coverage_values,
                             predicted_probability = round(predicted_probabilities, 2))

threshold <- 0.95

predicted_data$sequence <- ifelse(predicted_data$predicted_probability > threshold, "Yes", "No")

head(predicted_data)


x<- nextseq$library_conc_nm
y <- nextseq$cluster_density

predicted_probabilities <- prediction_sequence_generation(x, y, model)

predicted_data  <- data.frame(library_conc_nm = x, 
                              cluster_density = y, 
                              generated = nextseq$generated,
                              predicted_probability = round(predicted_probabilities, 4))

threshold <- 0.90

predicted_data$sequence <- ifelse(
  predicted_data$predicted_probability > threshold, "Yes", "No")
```

## Prediction Table.
```{r echo=FALSE, include=FALSE}
predicted_data %>%
  head() %>%
  gt() %>%
  cols_label(
    library_conc_nm = "Library Concentration",
    predicted_probability = "Predicted Probability",
    sequence = "Sequence Generated"
  ) %>%
  tab_header(
    title = md("**Predicted of Sequence Generation Summary**"),
    subtitle = md("*Showing the first few rows of predicted data*")
  ) %>%
  tab_spanner(
    label = "Prediction Information",
    columns = c(library_conc_nm, predicted_probability)
  ) %>%
  tab_spanner(
    label = "Sequence Information",
    columns = c(sequence)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold", color = "darkblue"),
    locations = cells_title()
  ) %>%
  tab_style(
    style = cell_text(weight = "normal", color ="black"),
    locations = cells_column_labels(columns = everything())
  ) %>%
 tab_style(
    style = cell_text(align = "center", color = "#25dc00"),
    locations = cells_body(rows = sequence == "Yes", columns = sequence)
  ) %>%
  tab_style(
    style = cell_text(align = "center", color = "#ff5d12"),
    locations = cells_body(rows = sequence == "No", columns = sequence)
  )
```

## Prediction Model for the probability of sequence generation based on library concentration and Cluster Density using a Random Forest model.
```{r message=FALSE, warning=FALSE, echo=FALSE}
library(caret)
set.seed(123)
train_control <- trainControl(method = "cv", number = 10, 
                              classProbs = TRUE, 
                              summaryFunction = twoClassSummary)
rf_model <- train(generated ~ library_conc_nm + cluster_density, 
                  data = nextseq,
                  method = "rf",
                  trControl = train_control,
                  metric = "ROC")


x <- nextseq$library_conc_nm
y <- nextseq$cluster_density

predict_sequence_generation <- function(library_conc, 
                                        cluster_density, 
                                        rf_model) {
  prob <- predict(rf_model, newdata = nextseq, type = "prob")$Yes
  return(prob)
}

predicted_probs <- predict_sequence_generation(x, y, rf_model)
predicted_data <- data.frame(library_conc_nm = x,
                             cluster_density = y,
                             predicted_probability = predicted_probs,
           generate = nextseq$sequence)

threshold <- 0.90

predicted_data$predicted <- ifelse(
  predicted_data$predicted_probability >= threshold, "Yes", "No")
# predicted_data %>% filter(!generate == "Yes" & predicted_probability >= 0.90)
```


## Prediction and sequence generation summary
```{r, echo=FALSE, include=FALSE}
predicted_data %>%
  filter(!generate == "Yes") %>%
  gt() %>%
  cols_label(
    library_conc_nm = "Library Concentration",
    predicted_probability = "Predicted Probability",
    predicted = "Predicted"
  ) %>%
  tab_header(
    title = md("**Predicted of Sequence Generation Summary**"),
    subtitle = md("*Showing the first few rows of predicted data*")
  ) %>%
  tab_spanner(
    label = "Prediction Information",
    columns = c(library_conc_nm, predicted_probability)
  ) %>%
  tab_spanner(
    label = "Sequence Information",
    columns = c(predicted)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold", color = "darkblue"),
    locations = cells_title()
  ) %>%
  tab_style(
    style = cell_text(weight = "normal", color ="black"),
    locations = cells_column_labels(columns = everything())
  ) %>%
 tab_style(
    style = cell_text(align = "center", color = "#25dc00"),
    locations = cells_body(rows = predicted == "Yes", columns = predicted)
  ) %>%
  tab_style(
    style = cell_text(align = "center", color = "#ff5d12"),
    locations = cells_body(rows = predicted == "No", columns = predicted)
  )
```

